Retrofitting Effect Handlers onto OCaml

### 2020-10-30 retro-concurrency.tex
Added semantics section

\newcommand{\oclos}[3]{\llparenthesis \olam{#1}{#2}, #3 \rrparenthesis}
\newcommand{\cclos}[3]{\llparenthesis \clam{#1}{#2}, #3 \rrparenthesis}
\newcommand{\clos}[3]{\llparenthesis \lam{#1}{#2}, #3 \rrparenthesis}
\newcommand{\kw}[1]{\text{\bf #1}}
\newcommand{\effval}[2]{\textsf{do} ~#1 ~#2}
\newcommand{\exnval}[1]{\textsf{exn} ~#1}

@@

The initial configuration for an expression $e$ is
$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$. We define {\sf continue} and
{\sf discontinue} as follows:

\begin{smathpar}
	\begin{array}{rcl}
		\textsf{continue} ~k ~e & = & k ~(\olam{x}{x}) ~e \\
		\textsf{discontinue} ~k ~l ~e & = & k ~(\olam{x}{\throw{l}{x}}) ~e
	\end{array}
\end{smathpar}

\begin{figure*}
	\begin{minipage}[t]{0.49\linewidth}
	\begin{smathpar}
	\begin{array}{rrcl}
		\text{Constants} & n & \coloneqq & ℤ \\
		\text{Abstractions} & \Lambda & \coloneqq & \lambda^o \mid \lambda^c \\
		\text{Expressions} & e & \coloneqq & n \mid x \mid e ~e \mid  \lam{x}{e} \mid e \odot e \mid \throw{l}{e} \\
											 &   & \mid      & \handle{e}{h} \mid \perform{l}{e} \\
		\text{Handlers} & h & \coloneqq & \{\caseval{x}{e}\} \mid \{\caseexn{l}{x}{e}\} \uplus h \\
		               &   & \mid & \{\caseeff{l}{x}{k}{e}\} \uplus h \\
		\text{Values} & v & \coloneqq & n \mid k \mid \clos{x}{e}{\env} \mid \effval{l}{k} \mid \exnval{l} \\
		\text{Frames} & r & \coloneqq & \farg{e}{\env} \mid \ffun{v} \mid \faritha{\odot}{e}{\env} \mid \farithb{\odot}{ℕ} \\
		\text{Environments} & \env & \coloneqq & \emptyset \mid \envext{\env}{x}{v} \\
	\end{array}
	\end{smathpar}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
	\begin{smathpar}
	\begin{array}{rrcl}
		\text{Handler Closures} & \hc & \coloneqq & (h,\env) \\
		\text{Frame List} & \fl & \coloneqq & [] \mid r :: \fl \\
		\text{Fibers} & \fiber & \coloneqq & (\fl, \hc) \\
		\text{Continuations} & k & \coloneqq & [] \mid \fiber \lhd k \\
		\text{C stacks} & \cstack & \coloneqq & \cstacka{\fl}{\ostack}\\
		\text{OCaml stacks} & \ostack & \coloneqq & \ostacka{k}{\cstack} \mid \ostackemp \\
		\text{Stacks} & \stack & \coloneqq & \cstack \mid \ostack \\
		\text{Terms} & \term & \coloneqq & e \mid v \\
		\text{Configurations} & \config & \coloneqq & \configa{\tau}{\env}{\stack}
	\end{array}
	\end{smathpar}
	\end{minipage}
	\caption{Term Syntax}
	\label{sem:syntax}
\end{figure*}

\begin{figure*}
	\begin{smathpar}
		\begin{array}{rrcl}
			\textsc{Var}     & (x,\env, \fl) & \rightsquigarrow
			                 & (\env(x), \env, \fl) \\
			\textsc{Arith1}  & (e_1 \odot e_2, \env, \fl) & \rightsquigarrow
			                 & (e_1,\env, \faritha{\odot}{e_2}{\env}::\fl) \\
			\textsc{Arith2}  & (n_1, \_, \faritha{\odot}{e_2}{\env}::\fl) & \rightsquigarrow
			                 & (e_2, \env, \farithb{\odot}{n_1}::\fl) \\
			\textsc{Arith3}  & (n_2, \env, \farithb{\odot}{n_1}::\fl) & \rightsquigarrow
			                 & (\llbracket n_1 \odot n_2 \rrbracket, \env, \fl) \\
			\textsc{App1}    & (e_1 ~e_2, \env, \fl) & \rightsquigarrow
			                 & (e_1, \env, \farg{e_2}{\env}::\fl) \\
			\textsc{App2}    & (\lam{x}{e}, \env, \fl) & \rightsquigarrow
			                 & (\clos{x}{e}{\env}, \env, \fl) \\
			\textsc{App3}    & (\clos{x}{e_1}{\env_1}, \_, \farg{e_2}{\env_2}::\fl) & \rightsquigarrow
			                 & (e_2, \env_2, \ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
			\textsc{Resume1} & (k,\_,\farg{e_1}{\env_1}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
			                 & (e_1, \env_1, \ffun{k}::\farg{e_2}{\env_2}::\fl) \\
			\textsc{Resume2} & (\clos{x}{e_1}{\env_1}, \_, \ffun{k}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
			                 & (e_2, \env_2, \ffun{k}::\ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
			\textsc{Perform} & (\perform{l}{e}, \env, \fl) & \rightsquigarrow
			                 & (e, \env, \ffun{\effval{l}{[[],(\{\caseval{x}{x}\},\emptyset)]}}::\fl) \\
			\textsc{Raise}   & (\throw{l}{e}, \env, \fl) & \rightsquigarrow
											 & (e, \env, \ffun{\exnval{l}}::\fl)
		\end{array}
	\end{smathpar}
	\caption{Local Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
	\label{sem:step}
\end{figure*}

\begin{figure*}
	\begin{smathpar}
		\begin{array}{rrcll}
			\textsc{LocalO}   & (\term,\env,(\fl,\hc) \kcons k,\cstack) & \ostep
												& \configa{\term'}{\env'}{\ostacka{(\fl',\hc) \kcons k}{\cstack}}
												& \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
			\textsc{RetC}   & (v, \_, [([],(h,\env))], \cstack) & \ostep
												& \configa{e}{\envext{\env}{x}{v}}{\cstack}
												& \text{if } \{\caseval{x}{e}\} \in h \\
			\textsc{RetFib} & (v, \_, ([],(h,\env)) \kcons k, \cstack) & \ostep
												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstack}}
												& \text{if } \{\caseval{x}{e}\} \in h \\
			\textsc{CallO}    & (v, \_, (\oclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{(\fl,\hc) \kcons k}{\cstack}} \\
			\textsc{ExtCall}  & (v, \_, (\cclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
												& \configa{e}{\envext{\env}{x}{v}}{\cstacka{[]}{\ostacka{(\fl,\hc) \kcons k}{\cstack}}} \\
			\textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k, \cstack) & \ostep
												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ (\fl,\hc) \kcons k'}{\cstack}} \\
			\textsc{Handle}   & (\handle{e}{h}, \env, k, \cstack) & \ostep
												& \configa{e}{\env}{\ostacka{([],(h,\env)) \kcons k}{\cstack}} \\
			\textsc{EffUnHn}  & (v, \_, [\ffun{\effval{l}{k}}::\fl,(h,\env)], \cstack) & \ostep
												& \configa{e}{\emptyset}{\ostacka{k ~@~ [(\fl,(h,\env))]}{\cstack}}
												& \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
									& & & & \text{and } {e = \throw{\textsf{unhandled}}{0}} \\
			\textsc{EffHn}    & (v, \_, (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons k', \cstack) & \ostep
												& \configa{e}{\env[r \mapsto k''][x \mapsto v]}{\ostacka{k'}{\cstack}}
												& \text{if } \{\caseeff{l}{x}{r}{e}\} \in h \\
									& & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
			\textsc{EffFwd}   & (v, \env', (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
												& \configa{v}{\env'}{\ostacka{(\ffun{\effval{l}{k''}}::\fl',\hc') \kcons k'}{\cstack}}
												& \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
								  & & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
			\textsc{ExFwdC}   & (v, \env, [\ffun{\exnval{l}}::\_,(h,\_)], \cstacka{\fl'}{\ostack}) & \ostep
												& \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl'}{\ostack}}
												& \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
			\textsc{ExHn}     & (v, \_, (\ffun{\exnval{l}}::\_,(h,\env)) \kcons k', \cstack) & \ostep
												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k'}{\cstack}}
												& \text{if } \{\caseexn{l}{x}{e}\} \in h \\
			\textsc{ExFwdFib} & (v, \env, (\ffun{\exnval{l}}::\_,(h,\_)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
												& \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl',\hc') \kcons k'}{\cstack}}
												& \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
		\end{array}
	\end{smathpar}
	\caption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
	\label{sem:ostep}
\end{figure*}

\begin{figure*}
	\begin{smathpar}
		\begin{array}{rrcll}
			\textsc{LocalC}   & (\term,\env,\fl,\ostack) & \cstep
												& \configa{\term'}{\env'}{\cstacka{\fl'}{\ostack}}
												& \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
			\textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
			                  & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
			\textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
												& \configa{e}{\envext{\env}{x}{v}}{\cstacka{\fl}{\ostack}} \\
			\textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep
												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstacka{\fl}{\ostack}}}
												& \text{if } k = [[],(\{\caseval{x}{x}\},\emptyset)] \\
			\textsc{ExnFwdO}  & (v, \env, \ffun{\exnval{l}}::\_, \ostacka{(\fl,\hc) \kcons k}{\cstack}) & \cstep
												& \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl,\hc) \kcons k}{\cstack}}
		\end{array}
	\end{smathpar}
	\caption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
	\label{sem:cstep}
\end{figure*}

\begin{figure*}
	\begin{mathpar}
		\begin{array}{rc}

			\textsc{StepC} & \inferrule
			                    {(\term, \env, \fl, \ostack) \cstep \config}
													{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} \\ \\
			\textsc{StepO} & \inferrule
													{(\term, \env, k, \cstack) \ostep \config}
													{\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
		\end{array}
	\end{mathpar}
	\caption{Top-level Reductions -- $\config \step \config$.}
	\label{sem:top}
\end{figure*}

>>>>

\newcommand{\oclos}[3]{\llparenthesis \olam{#1}{#2}, #3 \rrparenthesis}
\newcommand{\cclos}[3]{\llparenthesis \clam{#1}{#2}, #3 \rrparenthesis}
\newcommand{\clos}[3]{\llparenthesis \lam{#1}{#2}, #3 \rrparenthesis}
\newcommand{\kw}[1]{\text{\bf #1}}
\newcommand{\effval}[2]{\textsf{eff} ~#1 ~#2}
\newcommand{\exnval}[1]{\textsf{exn} ~#1}

@@

\begin{figure*}
	\begin{minipage}[t]{0.49\linewidth}
	\begin{smathpar}
	\begin{array}{rrcl}
		\text{Constants} & n & \coloneqq & ℤ \\
		\text{Abstractions} & \Lambda & \coloneqq & \lambda^o \mid \lambda^c \\
		\text{Expressions} & e & \coloneqq & n \mid x \mid e ~e \mid  \lam{x}{e} \mid e \odot e \mid \throw{l}{e} \\
											 &   & \mid      & \handle{e}{h} \mid \perform{l}{e} \\
		\text{Handlers} & h & \coloneqq & \{\caseval{x}{e}\} \mid \{\caseexn{l}{x}{e}\} \uplus h \\
		               &   & \mid & \{\caseeff{l}{x}{k}{e}\} \uplus h \\
		\text{Values} & v & \coloneqq & n \mid k \mid \clos{x}{e}{\env} \mid \effval{l}{k} \mid \exnval{l} \\
		\text{Frames} & r & \coloneqq & \farg{e}{\env} \mid \ffun{v} \mid \faritha{\odot}{e}{\env} \mid \farithb{\odot}{ℕ} \\
		\text{Environments} & \env & \coloneqq & \emptyset \mid \envext{\env}{x}{v} \\
	\end{array}
	\end{smathpar}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
	\begin{smathpar}
	\begin{array}{rrcl}
		\text{Handler Closures} & \hc & \coloneqq & (h,\env) \\
		\text{Frame List} & \fl & \coloneqq & [] \mid r :: \fl \\
		\text{Fibers} & \fiber & \coloneqq & (\fl, \hc) \\
		\text{Continuations} & k & \coloneqq & [] \mid \fiber \kcons k \\
		\text{C stacks} & \cstack & \coloneqq & \cstacka{\fl}{\ostack}\\
		\text{OCaml stacks} & \ostack & \coloneqq & \ostacka{k}{\cstack} \mid \ostackemp \\
		\text{Stacks} & \stack & \coloneqq & \cstack \mid \ostack \\
		\text{Terms} & \term & \coloneqq & e \mid v \\
		\text{Configurations} & \config & \coloneqq & \configa{\tau}{\env}{\stack}
	\end{array}
	\end{smathpar}
	\end{minipage}
	\caption{Syntax of expressions and configurations}
	\label{sem:syntax}
\end{figure*}

We present the operational semantics for a core language of effect handlers in
Multicore OCaml. The operational semantics intends to faithfully capture the
semantics of Multicore OCaml effect handlers, and to inform an implementation.
We will highlight the differences whenever we differ from the semantics of the
implementation. We have included an executable version of the dynamic semantics
as an OCaml program in the supplementary material \kc{include hl-semantics.ml}.

\subsubsection{Syntax}

Figure~\ref{sem:syntax} presents the syntax of expressions and configurations.
Our expressions consists of integer constants ($n$), variables ($x$),
abstraction ($\lam{x}{x}$), application ($e~e$), arithmetic expressions ($e
\odot e$) where $\odot$ ranges over $\{+,-,*,/\}$, raising exceptions
($\throw{l}{e}$), performing effects ($\perform{l}{e}$), and handling effects
($\handle{e}{h}$). Abstractions come in two forms: OCaml abstractions
($\lambda^o$) and C abstractions ($\lambda^c$). The handler consists of a
return case ($\caseval{x}{e}$), zero or more exception cases
($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and body $e$, and zero or
more effect cases ($\caseeff{l}{x}{k}{e}$) with label $l$, parameter $x$,
continuation $k$ and body $e$.

The operational semantics is an extension of a CEK machine
semantics~\cite{Felleisen} for effect handlers, following the abstract machine
semantics of Hillerstrom et al.~\cite{Hillerstrom2020}. The crucial difference
from Hillerstrom et al. is that our stacks are composed of alternating sequence
of OCaml and C stack segments. The program state is captured as configuration
$\config \coloneqq \configa{\tau}{\env}{\stack}$ with the current term $\tau$
under evaluation, its environment $\env$ and the current stack $\stack$. The
term is either an expression $e$ or a value $v$. The value are integer
constants $n$, continuations $k$, closure $\clos{x}{e}{\env}$, effect begin
performed $\effval{l}{k}$ and exception being raised $\exnval{l}$. The
environment is a map from variables to values.

The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
frames $\fl$, and, importantly, the OCaml stack $\ostack$ under it. Every OCaml
stack is either an empty stack $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$
with the C stack $\cstack$ under it. Thus the stack is an alternating sequence
of C and OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The
frame list $\fl$ is composed of individual frames $r$, which is one of an
argument frame $\farg{e}{\env}$ with the expression $e$ at the argument
position of an application with its environment $\env$, a function frame
$\ffun{v}$ with the value at the function position of an application, and
frames for evaluating the arguments of an arithmetic expression.

Continuations $k$ are either empty $[]$ or non-empty sequence of \emph{fibers}.
A fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler
closure $\hc \coloneqq (h,\env)$, which is a pair of handler $h$ and its
environment $\env$.

\subsubsection{Top-level reductions}

\begin{figure}
	\begin{mathpar}
		\begin{array}{rc}

			\textsc{StepC} & \inferrule
			                    {(\term, \env, \fl, \ostack) \cstep \config}
													{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} \\ \\
			\textsc{StepO} & \inferrule
													{(\term, \env, k, \cstack) \ostep \config}
													{\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
		\end{array}
	\end{mathpar}
	\caption{Top-level Reductions -- $\config \step \config$.}
	\label{sem:top}
\end{figure}

The initial configuration for an expression $e$ is
$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
stack are empty. The top-level reductions (Figure~\ref{sem:top}) can be
performed by either by taking a C step $\cstep$ or an OCaml step $\ostep$.
Intuitively, this splits the world into executing C code and OCaml code.

\begin{figure*}
	\begin{smathpar}
		\begin{array}{rrcll}
			\textsc{AdminC}   & (\term,\env,\fl,\ostack) & \cstep
												& \configa{\term'}{\env'}{\cstacka{\fl'}{\ostack}}
												& \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
			\textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
			                  & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
			\textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
												& \configa{e}{\envext{\env}{x}{v}}{\cstacka{\fl}{\ostack}} \\
			\textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep
												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstacka{\fl}{\ostack}}}
												& \text{if } k = [[],(\{\caseval{x}{x}\},\emptyset)] \\
			\textsc{ExnFwdO}  & (v, \env, \ffun{\exnval{l}}::\_, \ostacka{(\fl,\hc) \kcons k}{\cstack}) & \cstep
												& \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl,\hc) \kcons k}{\cstack}}
		\end{array}
	\end{smathpar}
	\caption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
	\label{sem:cstep}
\end{figure*}

\begin{figure*}
	\begin{smathpar}
		\begin{array}{rrcl}
			\textsc{Var}     & (x,\env, \fl) & \rightsquigarrow
			                 & (\env(x), \env, \fl) \\
			\textsc{Arith1}  & (e_1 \odot e_2, \env, \fl) & \rightsquigarrow
			                 & (e_1,\env, \faritha{\odot}{e_2}{\env}::\fl) \\
			\textsc{Arith2}  & (n_1, \_, \faritha{\odot}{e_2}{\env}::\fl) & \rightsquigarrow
			                 & (e_2, \env, \farithb{\odot}{n_1}::\fl) \\
			\textsc{Arith3}  & (n_2, \env, \farithb{\odot}{n_1}::\fl) & \rightsquigarrow
			                 & (\llbracket n_1 \odot n_2 \rrbracket, \env, \fl) \\
			\textsc{App1}    & (e_1 ~e_2, \env, \fl) & \rightsquigarrow
			                 & (e_1, \env, \farg{e_2}{\env}::\fl) \\
			\textsc{App2}    & (\lam{x}{e}, \env, \fl) & \rightsquigarrow
			                 & (\clos{x}{e}{\env}, \env, \fl) \\
			\textsc{App3}    & (\clos{x}{e_1}{\env_1}, \_, \farg{e_2}{\env_2}::\fl) & \rightsquigarrow
			                 & (e_2, \env_2, \ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
			\textsc{Resume1} & (k,\_,\farg{e_1}{\env_1}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
			                 & (e_1, \env_1, \ffun{k}::\farg{e_2}{\env_2}::\fl) \\
			\textsc{Resume2} & (\clos{x}{e_1}{\env_1}, \_, \ffun{k}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
			                 & (e_2, \env_2, \ffun{k}::\ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
			\textsc{Perform} & (\perform{l}{e}, \env, \fl) & \rightsquigarrow
			                 & (e, \env, \ffun{\effval{l}{[[],(\{\caseval{x}{x}\},\emptyset)]}}::\fl) \\
			\textsc{Raise}   & (\throw{l}{e}, \env, \fl) & \rightsquigarrow
											 & (e, \env, \ffun{\exnval{l}}::\fl)
		\end{array}
	\end{smathpar}
	\caption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
	\label{sem:step}
\end{figure*}

\subsubsection{C reductions}

The C reduction steps are given in Figure~\ref{sem:cstep}. We can take a C step
by taking an administrative reduction step $\rightsquigarrow$. The
administrative reductions are common to both C and OCaml. The rules
\textsc{Var}, \textsc{Arith1}, \textsc{Arith2}, \textsc{App1}, \textsc{App2}
and \textsc{App3} are standard. \textsc{Arith3} performs the arithmetic
operation on the integers ($\llbracket n_1 \odot n_2 \rrbracket$).
\textsc{Raise} pushes an function frame with exception value to indicate that
an exception is being raised. Similarly, \textsc{Perform} pushes a function
frame with an effect value with an empty continuation
$[[],(\{\caseval{x}{x}\},\emptyset)]$ with no captured frames and an empty
handler with an identity return case alone. \textsc{Resume1} and
\textsc{Resume2} are used for resuming continuations, and we shall defer their
explanation till later.

Continuing with the rest of the C reduction steps, \textsc{CallC} captures the
behaviour of calling a C function. Since the program is currently executing C,
indicated by the current stack being a C stack $\cstacka{\fl}{\ostack}$, we can
perform the call on the current stack. In case the abstraction is an OCaml
abstraction (\textsc{Callback}), we create an OCaml stack with the C stack as
its tail, with the current continuation being empty. This captures the
behaviour of calling back into OCaml from C. \textsc{RetToO} returns a value to
the enclosing OCaml stack. \textsc{ExnFwdO} forwards a raised exception to the
enclosing OCaml stack, ignoring the unwinding the rest of the frames. This
captures the semantics of raising OCaml exceptions from C.

\subsubsection{OCaml Reductions}

\begin{figure*}
	\begin{smathpar}
		\begin{array}{rrcll}
			\textsc{AdminO}   & (\term,\env,(\fl,\hc) \kcons k,\cstack) & \ostep
												& \configa{\term'}{\env'}{\ostacka{(\fl',\hc) \kcons k}{\cstack}}
												& \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
			\textsc{RetToC}   & (v, \_, [([],(h,\emptyset))], \cstack) & \ostep
												& \configa{v}{\env}{\cstack}
												& \text{if } h = \{\caseval{x}{x}\} \\
			\textsc{RetFib} & (v, \_, ([],(h,\env)) \kcons k, \cstack) & \ostep
												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstack}}
												& \text{if } \{\caseval{x}{e}\} \in h \\
			\textsc{CallO}    & (v, \_, (\oclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{(\fl,\hc) \kcons k}{\cstack}} \\
			\textsc{ExtCall}  & (v, \_, (\cclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep
												& \configa{e}{\envext{\env}{x}{v}}{\cstacka{[]}{\ostacka{(\fl,\hc) \kcons k}{\cstack}}} \\
			\textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k, \cstack) & \ostep
												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ ((\fl,\hc) \kcons k')}{\cstack}} \\
			\textsc{Handle}   & (\handle{e}{h}, \env, k, \cstack) & \ostep
												& \configa{e}{\env}{\ostacka{([],(h,\env)) \kcons k}{\cstack}} \\
			\textsc{EffUnHn}  & (v, \_, [\ffun{\effval{l}{k}}::\fl,(h,\env)], \cstack) & \ostep
												& \configa{e}{\emptyset}{\ostacka{k ~@~ [(\fl,(h,\env))]}{\cstack}}
												& \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
									& & & & \text{and } {e = \throw{\textsf{unhandled}}{0}} \\
			\textsc{EffHn}    & (v, \_, (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons k', \cstack) & \ostep
												& \configa{e}{\env[r \mapsto k''][x \mapsto v]}{\ostacka{k'}{\cstack}}
												& \text{if } \{\caseeff{l}{x}{r}{e}\} \in h \\
									& & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
			\textsc{EffFwd}   & (v, \env', (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
												& \configa{v}{\env'}{\ostacka{(\ffun{\effval{l}{k''}}::\fl',\hc') \kcons k'}{\cstack}}
												& \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
								  & & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
			\textsc{ExFwdC}   & (v, \env, [\ffun{\exnval{l}}::\_,(h,\_)], \cstacka{\fl'}{\ostack}) & \ostep
												& \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl'}{\ostack}}
												& \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
			\textsc{ExHn}     & (v, \_, (\ffun{\exnval{l}}::\_,(h,\env)) \kcons k', \cstack) & \ostep
												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k'}{\cstack}}
												& \text{if } \{\caseexn{l}{x}{e}\} \in h \\
			\textsc{ExFwdFib} & (v, \env, (\ffun{\exnval{l}}::\_,(h,\_)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
												& \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl',\hc') \kcons k'}{\cstack}}
												& \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
		\end{array}
	\end{smathpar}
	\caption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
	\label{sem:ostep}
\end{figure*}

OCaml reductions are given in Figure~\ref{sem:ostep}. In OCaml, reductions
always occur on the top-most fiber in the current stack. \textsc{AdminO}
performs administrative reductions. \textsc{CallO} evaluates an OCaml function
on the current stack. \textsc{ExtCall} captures the behaviour of external
calls. On an external call, we create a C stack with no frames, whose tail is
the current OCaml stack, and evaluate the C functions body in the new stack.
\textsc{RetToC} returns a value to the enclosing C stack. We have exactly one
fiber on the stack, and this was created in the rule \textsc{Callback}, whose
handler has identity return case alone and the environment is empty.
\textsc{RetFib} returns value from a fiber to the previous one, evaluating the
body of the return case.

The rule \textsc{Handle} installs a handler by pushing a fiber with no frames
and the given handler. The rule \textsc{ExHn} handles an exception, if the
current handler has a matching exception case. Observe that fiber on which the
exception was raise has been unwound. The rule \textsc{ExFwdC} forwards the
exception to C. Here, there is exactly one fiber on the current stack, and the
handler does not have a matching exception case. The rule \textsc{ExFwdFib}
forwards the exception to the next fiber if the current handler does not handle
it.

The rule \textsc{EffHn} captures the handling of effects when the current
handler has a matching effect case. We evaluate the body of the matching case,
and bind the continuation parameter |r| to the captured continuation. Observe
that the captured continuation |k''| includes the current handler. Intuitively,
the handler wraps around captured continuation. This gives Multicore OCaml
effect handlers deep handler semantics~\cite{}. \textsc{EffFwd} forwards the
effect to the outer fiber, and extends the captured continuation |k''| in the
process. Recall that we do not capture C frames as part of a continuation.
\textsc{EffUnHn} models unhandled effect. If the effect bubbles up to the top
fiber, which we know does not have an effect case (see \textsc{Callback} rule),
we raise \textsc{unhandled} exception at the point where the corresponding
effect was performed. This is achieved by appending the captured continuation
to the front of the current continuation.

Observe that |continue| and |discontinue| are not part of the expressions. They
are encoded as follows:
\begin{smathpar}
	\begin{array}{rcl}
		\textsf{continue} ~k ~e & = & k ~(\olam{x}{x}) ~e \\
		\textsf{discontinue} ~k ~l ~e & = & k ~(\olam{x}{\throw{l}{x}}) ~e
	\end{array}
\end{smathpar}

Intuitively, resuming a continuation in both the cases involves evaluating the
appropriate abstraction on top of the continuation. We perform the
administrative reductions \textsc{Resume1} and \textsc{Resume2} to evaluate the
arguments to |continue| and |discontinue|. The rule \textsc{Resume} appends the
given continuation to the front of the current continuation, and evaluates the
body of the closure.
